---
title: "JIVE Integration (Dataset 2)"
author: "Joey Hastings"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
    thumbnails: FALSE
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.dim = c(9, 6)
  )
```

# Load Packages

```{r load_packages, message = FALSE}
library(tictoc)
library(tidyverse)
library(cowplot)
library(ggExtra)
library(ggthemes)

library(splatter)
library(scater)
library(scran)
library(r.jive)
library(Seurat)
library(factoextra)
library(SLIDE)
library(RSpectra)
library(scRNAseq)
library(lme4)

library(kBET) # kBET
library(cluster) # ASW
library(lisi) # LISI
library(mclust) # ARI

source("PVCA.R")

theme_set(theme_few())

method_name <- "JIVE"
dataset_name <- "dataset2"
```

# Bacher T-Cell Data

## Load Data from `scRNAseq` Package

* Remove low count cell types 

```{r load_data}
BacherTCellData <- BacherTCellData()
table(BacherTCellData$batch)
table(BacherTCellData$batch, BacherTCellData$new_cluster_names)

# BacherTCellData <- BacherTCellData[, !(BacherTCellData$new_cluster_names %in% c("Cycling", "Type-1 IFN signature"))]
# table(BacherTCellData$batch)
# table(BacherTCellData$batch, BacherTCellData$new_cluster_names)

bacher_metadata <- as.data.frame(colData(BacherTCellData))
bacher_seurat <- CreateSeuratObject(counts(BacherTCellData), meta.data = bacher_metadata)

#################
# Identify HVGs #
#################

# split the dataset into a list of seurat objects (one for each batch)
bacher_seurat_batch <- SplitObject(bacher_seurat, split.by = "batch")

# normalize and identify variable features for each dataset independently
bacher_seurat_batch <- lapply(X = bacher_seurat_batch, FUN = function(x) {
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = bacher_seurat_batch)

########################################################

# Select only HVGs
BacherTCellData <- BacherTCellData[features, ]

# Subset to three batches
BacherTCellData <- BacherTCellData[, colData(BacherTCellData)$batch %in% c(14, 15)]

# Final SCE object
data_SCE <- BacherTCellData
colData(data_SCE)$Batch <- factor(colData(data_SCE)$batch)
colData(data_SCE)$Cluster <- factor(colData(data_SCE)$new_cluster_names)
```

```{r raw_viz, message = FALSE}
data_SCE <- logNormCounts(data_SCE)

# Dimension reduction visualization
set.seed(1)
data_SCE <- runPCA(data_SCE, ncomponents = 30, ntop = 2000)
data_SCE <- runTSNE(data_SCE)
data_SCE <- runUMAP(data_SCE)

orig_title <- labs(title = "Original Data")

orig_p1 <- plot_grid(
  plotPCA(data_SCE, color_by = "Batch")   + scale_color_fivethirtyeight(name = "Batch") +
    orig_title,
  plotPCA(data_SCE, color_by = "Cluster") + scale_color_gdocs(name = "Cluster"),
  nrow=2, align = "v"
  ) + labs(title = "Test")

orig_p2 <- plot_grid(
  plotTSNE(data_SCE, color_by = "Batch")   + scale_color_fivethirtyeight(name = "Batch") +
    orig_title,
  plotTSNE(data_SCE, color_by = "Cluster") + scale_color_gdocs(name = "Cluster"),
  nrow=2, align = "v"
  )

orig_p3 <- plot_grid(
  plotUMAP(data_SCE, color_by = "Batch")   + scale_color_fivethirtyeight(name = "Batch") +
    orig_title,
  plotUMAP(data_SCE, color_by = "Cluster") + scale_color_gdocs(name = "Cluster"),
  nrow=2, align = "v"
  )

orig_p1 # ggsave("orig_pca.png", orig_p1)
orig_p2 # ggsave("orig_tsne.png", orig_p2)
orig_p3 # ggsave("orig_umap.png", orig_p3)
```

## Separate Batches

```{r batches_clusters}
batches <- data_SCE$Batch
# Frequency of batches in simulation
table(batches)

clusters <- data_SCE$Cluster
# Frequency of cell types in simulation
table(clusters)

rawcounts <- as.matrix(counts(data_SCE))
dim(rawcounts)
```

## PVCA

```{r sim1_pvca}
# Perform PVCA for 10 random samples of size 1000 (more computationally efficient)
pvca.res <- matrix(nrow = 10, ncol = 3)
meta <- cbind(batch = batches, cluster = clusters)

for (i in 1:10){
  sample <- sample(1:ncol(rawcounts), 1000, replace = FALSE)
  pvca.res[i, ] <- PVCA(rawcounts[, sample], meta[sample, ], threshold = 0.6, inter = FALSE)
}

# Average effect size across samples
pvca.means <- colMeans(pvca.res)
names(pvca.means) <- c(colnames(meta), "resid")

# Plot PVCA
pvca.plot <- PlotPVCA(pvca.means, "PVCA of Raw Count Data")
pvca.plot
```

## JIVE

```{r, eval = FALSE}
n_batches <- length(unique(batches))
unq_batches <- sort(unique(batches))

jive_data <- NULL
all_clusters <- NULL

for (i in 1:n_batches) {
  jive_data[[unq_batches[i]]] <- t(rawcounts[, batches == unq_batches[i]])
  all_clusters[[unq_batches[i]]] <- clusters[batches == unq_batches[i]]
}

CORES <- parallel::detectCores()

tic("JIVE v2 runtime (given)")
JIVE_results <- jive_v2(jive_data, rankJ = 5, rankA = rep(10, length(jive_data)), method = "given", maxiter = 10000, CORES = CORES)
# JIVE v2 runtime (given): 789.33 sec elapsed
toc()

saveRDS(JIVE_results, file = "data/JIVE_v2_dataset2_b14_b15.rds")
```


Summary of estimated ranks and variance explained by joint/individual/residual for each approach:

```{r jive1_summary}
JIVE_results <- readRDS(file = "data/JIVE_v2_dataset2_b14_b15.rds")
summary(JIVE_results)
plot(JIVE_results)
```

### Visualization using `SCE`

```{r jive_viz_setup}
joint <- do.call(rbind, JIVE_results$joint)
indiv <- do.call(rbind, JIVE_results$individual)

joint_svd <- svds(joint, k = JIVE_results$rankJ)

var_explained <- data.frame(
  sv = joint_svd$d,
  cum_sv = cumsum(joint_svd$d),
  tot_sv = sum(joint_svd$d)
  ) %>%
  mutate(
    pct_sv = sv / tot_sv,
    cumpct_sv = cum_sv / tot_sv,
    PC = paste0("PC", row_number())
  )

var_explained %>%
  ggplot(aes(x = PC)) +
  geom_col(aes(y = cumpct_sv, fill = PC), color = "black") +
  geom_col(aes(y = cumpct_sv - pct_sv), fill = "gray", color = "black") +
  geom_label(aes(y = cumpct_sv - (pct_sv / 2), label = paste0(round(pct_sv * 100, 2), "%"))) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Principal Component",
    y = "Percentage",
    title = "Percentage of Total Variance Explained"
  )

integrated_final <- joint
```

```{r jive_viz1}
integrated_SCE <- SingleCellExperiment(list(logcounts = t(integrated_final)))
colData(integrated_SCE)$Batch <- batches
colData(integrated_SCE)$Cluster <- clusters

set.seed(1)
# Original PCA based on joint matrices for each batch
integrated_SCE <- runPCA(integrated_SCE, ncomponents = 30, ntop = 2000)

# t-SNE based on original PC
integrated_SCE <- runTSNE(integrated_SCE)

# UMAP based on original PC
integrated_SCE <- runUMAP(integrated_SCE)

title_caption <-
  labs(
    title = "Integrated Data",
    subtitle = paste0(
      "Joint Rank: ",
      JIVE_results$rankJ,
      ", Individual Ranks: ",
      paste(JIVE_results$rankA, collapse = ", ")
    )
  )

integrated_p1 <- plot_grid(
  plotPCA(integrated_SCE, color_by = "Batch") +
  scale_color_fivethirtyeight(name = "Batch") +
    title_caption,
  plotPCA(integrated_SCE, color_by = "Cluster") + scale_color_gdocs(name = "Cluster"),
  nrow=2, align = "v"
  )

integrated_p2 <- plot_grid(
  plotTSNE(integrated_SCE, color_by = "Batch") +
  scale_color_fivethirtyeight(name = "Batch") +
    title_caption,
  plotTSNE(integrated_SCE, color_by = "Cluster") + scale_color_gdocs(name = "Cluster"),
  nrow=2, align = "v"
  )

integrated_p3 <- plot_grid(
  plotUMAP(integrated_SCE, color_by = "Batch") +
    scale_color_fivethirtyeight(name = "Batch") +
    title_caption,
  plotUMAP(integrated_SCE, color_by = "Cluster") + scale_color_gdocs(name = "Cluster"),
  nrow=2, align = "v"
  )

integrated_p1 # ggsave(paste0("dataset2_j", JIVE_results$rankJ, "_a_15_14_integrated_pca.png"), integrated_p1)
integrated_p2 # ggsave(paste0("dataset2_j", JIVE_results$rankJ, "_a_15_14_integrated_tsne.png"), integrated_p2)
integrated_p3 # ggsave(paste0("dataset2_j", JIVE_results$rankJ, "_a_15_14_integrated_umap.png"), integrated_p3)
```

```{r jive1_clust1, eval = FALSE, echo = FALSE}
##############
# Clustering #
##############

clusters_PCA <- clusterCells(
  integrated_SCE,
  use.dimred = "PCA",
  BLUSPARAM = bluster::SNNGraphParam(
    cluster.fun = "louvain",
    cluster.args = list(resolution = 0.5)
    )
  )
summary(clusters_PCA)

###

clusters_TSNE <- clusterCells(
  integrated_SCE,
  use.dimred = "TSNE",
  BLUSPARAM = bluster::SNNGraphParam(
    cluster.fun = "louvain", 
    cluster.args = list(resolution = 0.5)
    )
  )
summary(clusters_TSNE)

###

clusters_UMAP <- clusterCells(
  integrated_SCE,
  use.dimred = "UMAP",
  BLUSPARAM = bluster::SNNGraphParam(
    cluster.fun = "louvain", 
    cluster.args = list(resolution = 0.5)
    )
  )
summary(clusters_UMAP)

###

colData(integrated_SCE)$Cluster_PCA <- clusters_PCA
colData(integrated_SCE)$Cluster_TSNE <- clusters_TSNE
colData(integrated_SCE)$Cluster_UMAP <- clusters_UMAP

# Choose clusters
clusters <- colData(integrated_SCE)$Cluster_PCA

plotReducedDim(integrated_SCE, dimred = "PCA", color_by = "Cluster_PCA", shape_by = "Cluster")
plotReducedDim(integrated_SCE, dimred = "TSNE", color_by = "Cluster_TSNE", shape_by = "Cluster")
plotReducedDim(integrated_SCE, dimred = "UMAP", color_by = "Cluster_UMAP", shape_by = "Cluster")
```

### Calculate kBET

* Sub-sample data at varying percentages
  * 5%, 10%, 15%, 20%, 25%
  * Stratified sampling so that each batch has the same number of cells selected
    * E.g., assume 2 batches, sampling 10% of 1000 cells = 100 cells: sample 50 cells from each batch
* Calculate kBET rejection rates at each percentage
  * **Lower** RR indicate well-mixed batches
  * **Higher** RR indicate poorly-mixed batches
  
```{r kbet1}
# data: a matrix (rows: samples, columns: features (genes))
data <- svds(integrated_final, k = 30)$u

# batch: vector or factor with batch label of each cell 
batch <- batches

sample_size <- seq(0.05, 0.25, by = 0.05)
rejection_rate <- list()

set.seed(1)
for (i in 1:length(sample_size)) {
  k <- floor(nrow(data) * sample_size[i])
  
  cat("\nSample Size: ", sample_size[i], ", k = ", k, "\n", sep = "")
  cat("------------------------------\n")
 
  batch.estimate <- kBET(data, batch, k0 = k, plot = F, do.pca = F, verbose = T)
  
  cat("\n------------------------------\n")
  
  if (class(batch.estimate) == "list") {
    rejection_rate[[i]] <- batch.estimate$summary %>%
      rownames_to_column(var = "statistic") %>%
      mutate(sample_size = sample_size[i]) %>%
      select(sample_size, statistic, kBET_rr = kBET.observed)
  }
}

all_rr <- bind_rows(rejection_rate) %>%
  mutate(method = method_name) %>%
  pivot_wider(id_cols = c(method, sample_size), names_from = statistic, values_from = kBET_rr) %>%
  select(method, sample_size, median_rr = `50%`)

all_rr %>% write_csv(file = paste0("output/", method_name, "_", dataset_name, "_kBET.csv"))

all_rr %>%
  ggplot(aes(x = sample_size, y = median_rr)) + 
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(labels = scales::percent) +
  expand_limits(y = c(0, 1)) + 
  labs(
    x = "% sample size",
    y = "kBET (rejection rate)",
    color = "Cell Group",
    title = "kBET Rejection Rates",
    subtitle = method_name
  )

all_rr %>%
  ggplot(aes(x = sample_size, y = I(1 - median_rr))) + 
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(labels = scales::percent) +
  expand_limits(y = c(0, 1)) + 
  labs(
    x = "% sample size",
    y = "kBET (acceptance rate)",
    color = "Cell Group",
    title = "kBET Acceptance Rates",
    subtitle = method_name
  )
```

### Calculate Average Silhouette Width

Original batch label and cell types/groups will be considered the clusters to compute $ASW_{batch}$ and $ASW_{group}$

* Compute average silhouette width (ASW) for the first 30 PCs of 80% sub-sample of joint data
  * ASW ranges between -1 and 1
  * Values **close to 1** indicate the cell is well-matched to its cluster (i.e., batch label)
  * Values **close to -1** indicate the cell is **not** well-matched to its cluster (i.e., batch label)
    * We hope to see lower values for $ASW_{batch}$ since it is indicative of well-mixed batches
    * We hope to see higher values for $ASW_{group}$ since it is indicative that distinct cell groups were preserved after batch-mixing
* Repeat `20` times
* Calculate the median ASW from the `20` runs to ensure stability of the measurement

```{r asw}
asw_repeat <- 20
npcs <- 30

asw_batch <- list()
asw_cluster <- list()
subset_prop <- 0.8 # subsample to 80% of the data
subset_size_total <- floor(length(batch) * subset_prop)

set.seed(1)
for (i in 1:asw_repeat) {
  subset_index <- 
    data.frame(index = 1:length(batch) , batch = batch) %>%
    slice_sample(n = subset_size_total, replace = F)
  
  subset_id <- subset_index %>%
    pull(index)
  
  data_pc <- svds(integrated_final[subset_id, ], k = npcs)$u
  
  dissimilarity_matrix <- daisy(data_pc)
  
  asw_batch_i <- silhouette(as.integer(factor(batches[subset_id])), dissimilarity_matrix)
  asw_batch[[i]] <- as.data.frame(rbind(summary(asw_batch_i)[["si.summary"]]))
  
  asw_cluster_i <- silhouette(as.integer(factor(clusters[subset_id])), dissimilarity_matrix)
  asw_cluster[[i]] <- as.data.frame(rbind(summary(asw_cluster_i)[["si.summary"]]))
}

all_asw_batch <- bind_rows(asw_batch) %>%
  mutate(Cluster = "Batch")

# ASW_batch summary
summary(all_asw_batch$Mean)

all_asw_cluster <- bind_rows(asw_cluster) %>%
  mutate(Cluster = "Cluster")

#ASW_group summary
summary(all_asw_cluster$Mean)

all_asw <- bind_rows(all_asw_batch, all_asw_cluster)

asw_plot <- ggplot(data = all_asw, aes(x = Cluster, y = Mean, fill = Cluster)) +
  geom_violin() +
  geom_boxplot(width = 0.2) +
  labs(x = "") +
  theme(legend.position = "none") +
  scale_fill_few()

asw_plot

all_asw %>%
  group_by(Cluster) %>%
  mutate(Method = method_name) %>%
  write_csv(file = paste0("output/", method_name, "_", dataset_name, "_ASW.csv"))
```

### Calculate LISI Scores

```{r jive_lisi}
# Performed for all cells

lisi <- compute_lisi(integrated_final, data.frame(batch = batches, cluster = clusters), c("batch", "cluster"), perplexity = 40) %>%
  pivot_longer(cols = c("batch", "cluster"), names_to = "type", values_to = "lisi")

# Looking for values close to the number of batches (2)
# Looking for values close to 1

ggplot(data = lisi, aes(x = type, y = lisi, fill = type)) +
  geom_violin() +
  geom_boxplot(width = 0.05)
```

<!-- ### Calculate Adjusted Rand Index -->

<!-- * Compute adjusted rand index (ARI) for a 80% sub-sample of clusters -->
<!--   * ARI is bounded above by 1 (i.e., perfect agreement between two partitions) -->
<!--   * Values **close to 1** indicate the two partition labels are similar -->
<!--   * Values **close to 0** indicate the two partition labels are not similar -->
<!--     * We hope to see lower values for $ARI_{batch}$ since it is indicative of well-mixed batches -->
<!--     * We hope to see higher values for $ARI_{group}$ since it is indicative that distinct cell groups were preserved after batch-mixing -->
<!-- * Repeat `20` times -->
<!-- * Calculate the median ARI from the `20` runs to ensure stability of the measurement -->

<!-- ```{r jive_ari} -->
<!-- ari_repeat <- 20 -->
<!-- npcs <- 30 -->

<!-- ari_batch <- list() -->
<!-- ari_group <- list() -->
<!-- subset_prop <- 0.8 # subsample to 80% of the data -->
<!-- subset_size_total <- floor(length(batch) * subset_prop) -->

<!-- set.seed(1) -->
<!-- for (i in 1:ari_repeat) { -->
<!--   subset_index <-  -->
<!--     data.frame(index = 1:length(batch) , batch = batch) %>% -->
<!--     slice_sample(n = subset_size_total, replace = F) -->

<!--   subset_id <- subset_index %>% -->
<!--     pull(index) -->

<!--   clusters <- kmeans(svd(integrated_final[subset_id, ])$u, centers = length(unique(groups)))$cluster -->

<!--   ari_batch[[i]] <- adjustedRandIndex(factor(batches[subset_id]), clusters[subset_id]) -->
<!--   ari_group[[i]] <- adjustedRandIndex(factor(groups[subset_id]), clusters[subset_id]) -->
<!-- } -->

<!-- all_ari_batch <- data.frame(ari = unlist(ari_batch)) %>% -->
<!--   mutate(cluster = "Batch") -->

<!-- # Lower values indicate well-mixed batches -->
<!-- summary(all_ari_batch$ari) -->

<!-- all_ari_group <- data.frame(ari = unlist(ari_group)) %>% -->
<!--   mutate(cluster = "Group")  -->

<!-- # Higher values indicate cell types were well preserved -->
<!-- summary(all_ari_group$ari) -->

<!-- all_ari <- bind_rows(all_ari_batch, all_ari_group) -->

<!-- ari_plot <- ggplot(data = all_ari, aes(x = cluster, y = ari, fill = cluster)) + -->
<!--   geom_violin() + -->
<!--   geom_boxplot(width = 0.2) -->

<!-- ari_plot -->
<!-- ``` -->

# Session Information

```{r session_info}
sessionInfo()
```
