---
title: "Seurat Integration (Dataset 2)"
author: "Joey Hastings"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
    thumbnails: FALSE
    toc_depth: 3
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.dim = c(9, 6)
  )
```

# Load Packages

```{r, message = FALSE}
library(tictoc)
library(tidyverse)
library(cowplot)
library(ggExtra)
library(ggthemes)

library(splatter)
library(scater)
library(scran)
library(r.jive)
library(Seurat)
library(factoextra)
library(SLIDE)
library(RSpectra)
library(scRNAseq)
library(lme4)

library(kBET) # kBET
library(cluster) # ASW
library(lisi) # LISI
library(mclust) # ARI

theme_set(theme_few())

method_name <- "Seurat"
dataset_name <- "dataset2"
```

# Bacher T-Cell Data

## Load Data from `scRNAseq` Package

```{r}
BacherTCellData <- BacherTCellData()
table(BacherTCellData$batch)
table(BacherTCellData$batch, BacherTCellData$new_cluster_names)
```

Perform cell batch integration as in their tutorial.

# Seurat

### Setup

* Create Seurat objects for each batch
  * Assign metadata to each object (batch number, cell type)
* Normalize the cell counts
* Identify most variable features (genes) 
* Choose most variable features from both batches for use in integration

```{r}
bacher_metadata <- as.data.frame(colData(BacherTCellData))
bacher_seurat <- CreateSeuratObject(counts(BacherTCellData), meta.data = bacher_metadata)

#################
# Identify HVGs #
#################

# split the dataset into a list of seurat objects (one for each batch)
bacher_seurat_batch <- SplitObject(bacher_seurat, split.by = "batch")

# normalize and identify variable features for each dataset independently
bacher_seurat_batch <- lapply(X = bacher_seurat_batch, FUN = function(x) {
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = bacher_seurat_batch)

########################################################

batch_indices <- which(unique(bacher_seurat$batch) %in% c(14, 15)) # subset batches
bacher_seurat_batch <- bacher_seurat_batch[batch_indices]
```

### Integration

* Perform integration by finding a set of anchors from the two batches (via CCA)
* Use the anchor list to integrate the two batches

```{r}
anchors <- FindIntegrationAnchors(object.list = bacher_seurat_batch, anchor.features = features)
combined <- IntegrateData(anchorset = anchors)
GetAssay(combined, "integrated")
```

### Integrated Analysis

* Integrated/combined data is centered/scaled
* First 30 PCs are computed/stored
* PC1-PC30 are used to perform t-SNE and UMAP dimensionality reduction
* Calculates k-nearest neighbors (`k = 20` by default) from PC1-PC30
* Identify clusters by using Shared Nearest Neighbors (SNN):
  * "First calculate k-nearest neighbors and construct the SNN graph"
  * "Then optimize the modularity function to determine clusters"

```{r}
# specify that we will perform downstream analysis on the corrected data note that the
# original unmodified data still resides in the 'RNA' assay
DefaultAssay(combined) <- "integrated"

# Run the standard workflow for visualization and clustering
set.seed(1)
combined <- ScaleData(combined, verbose = FALSE)
combined <- RunPCA(combined, npcs = 30, verbose = FALSE)
combined <- RunTSNE(combined, reduction = "pca", dims = 1:30)
combined <- RunUMAP(combined, reduction = "pca", dims = 1:30)
combined <- FindNeighbors(combined, reduction = "pca", dims = 1:30)
combined <- FindClusters(combined, resolution = 0.5)

#
combined <- AddMetaData(combined, factor(combined$batch), col.name = "Batch")
combined <- AddMetaData(combined, factor(combined$new_cluster_names), col.name = "Cluster")
```

Visualize integrated batches and clusters using UMAP:

```{r}
title_caption <-
  labs(
    title = "Integrated Data",
    subtitle = method_name
  )

integrated_p1 <- plot_grid(
  PCAPlot(combined, group.by = "Batch") +
  scale_color_fivethirtyeight(name = "Batch") +
    title_caption,
  PCAPlot(combined, group.by = "Cluster") + scale_color_gdocs(name = "Cluster") + labs(title = ""),
  nrow=2, align = "v"
  )

integrated_p2 <- plot_grid(
  TSNEPlot(combined, group.by = "Batch") +
  scale_color_fivethirtyeight(name = "Batch") +
    title_caption,
  TSNEPlot(combined, group.by = "Cluster") + scale_color_gdocs(name = "Cluster") + labs(title = ""),
  nrow=2, align = "v"
  )

integrated_p3 <- plot_grid(
  UMAPPlot(combined, group.by = "Batch") +
    scale_color_fivethirtyeight(name = "Batch") +
    title_caption,
  UMAPPlot(combined, group.by = "Cluster") + scale_color_gdocs(name = "Cluster") + labs(title = ""),
  nrow=2, align = "v"
  )

integrated_p1 # ggsave(paste0("dataset2_j", JIVE_results$rankJ, "_a_15_14_integrated_pca.png"), integrated_p1)
integrated_p2 # ggsave(paste0("dataset2_j", JIVE_results$rankJ, "_a_15_14_integrated_tsne.png"), integrated_p2)
integrated_p3 # ggsave(paste0("dataset2_j", JIVE_results$rankJ, "_a_15_14_integrated_umap.png"), integrated_p3)
```

```{r}
integrated_final <- t(combined@assays$integrated@scale.data)
batches <- combined$Batch
clusters <- combined$Cluster
```

### Calculate kBET

* Sub-sample data at varying percentages
  * 5%, 10%, 15%, 20%, 25%
  * Stratified sampling so that each batch has the same number of cells selected
    * E.g., assume 2 batches, sampling 10% of 1000 cells = 100 cells: sample 50 cells from each batch
* Calculate kBET rejection rates at each percentage
  * **Lower** RR indicate well-mixed batches
  * **Higher** RR indicate poorly-mixed batches
  
```{r}
# data: a matrix (rows: samples, columns: features (genes))
# data <- svds(integrated_final, k = 30)$u
data <- combined@reductions$pca@cell.embeddings

# batch: vector or factor with batch label of each cell 
batch <- combined$Batch

sample_size <- seq(0.05, 0.25, by = 0.05)
rejection_rate <- list()

set.seed(1)
for (i in 1:length(sample_size)) {
  k <- floor(nrow(data) * sample_size[i])
  
  cat("\nSample Size: ", sample_size[i], ", k = ", k, "\n", sep = "")
  cat("------------------------------\n")
 
  batch.estimate <- kBET(data, batch, k0 = k, plot = F, do.pca = F, verbose = T)
  
  cat("\n------------------------------\n")
  
  if (class(batch.estimate) == "list") {
    rejection_rate[[i]] <- batch.estimate$summary %>%
      rownames_to_column(var = "statistic") %>%
      mutate(sample_size = sample_size[i]) %>%
      select(sample_size, statistic, kBET_rr = kBET.observed)
  }
}

all_rr <- bind_rows(rejection_rate) %>%
  mutate(method = method_name) %>%
  pivot_wider(id_cols = c(method, sample_size), names_from = statistic, values_from = kBET_rr) %>%
  select(method, sample_size, median_rr = `50%`)

all_rr %>% write_csv(file = paste0("output/", method_name, "_", dataset_name, "_kBET.csv"))

all_rr %>%
  ggplot(aes(x = sample_size, y = median_rr)) + 
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(labels = scales::percent) +
  expand_limits(y = c(0, 1)) + 
  labs(
    x = "% sample size",
    y = "kBET (rejection rate)",
    color = "Cell Group",
    title = "kBET Rejection Rates",
    subtitle = method_name
  )

all_rr %>%
  ggplot(aes(x = sample_size, y = I(1 - median_rr))) + 
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(labels = scales::percent) +
  expand_limits(y = c(0, 1)) + 
  labs(
    x = "% sample size",
    y = "kBET (acceptance rate)",
    color = "Cell Group",
    title = "kBET Acceptance Rates",
    subtitle = method_name
  )
```

### Calculate Average Silhouette Width

Original batch label and cell types/groups will be considered the clusters to compute $ASW_{batch}$ and $ASW_{group}$

* Compute average silhouette width (ASW) for the first 30 PCs of 80% sub-sample of joint data
  * ASW ranges between -1 and 1
  * Values **close to 1** indicate the cell is well-matched to its cluster (i.e., batch label)
  * Values **close to -1** indicate the cell is **not** well-matched to its cluster (i.e., batch label)
    * We hope to see lower values for $ASW_{batch}$ since it is indicative of well-mixed batches
    * We hope to see higher values for $ASW_{group}$ since it is indicative that distinct cell groups were preserved after batch-mixing
* Repeat `20` times
* Calculate the median ASW from the `20` runs to ensure stability of the measurement

```{r}
asw_repeat <- 20
npcs <- 30

asw_batch <- list()
asw_cluster <- list()
subset_prop <- 0.8 # subsample to 80% of the data
subset_size_total <- floor(length(batch) * subset_prop)

set.seed(1)
for (i in 1:asw_repeat) {
  subset_index <- 
    data.frame(index = 1:length(batch) , batch = batch) %>%
    slice_sample(n = subset_size_total, replace = F)
  
  subset_id <- subset_index %>%
    pull(index)
  
  data_pc <- svds(integrated_final[subset_id, ], k = npcs)$u
  
  dissimilarity_matrix <- daisy(data_pc)
  
  asw_batch_i <- silhouette(as.integer(factor(batches[subset_id])), dissimilarity_matrix)
  asw_batch[[i]] <- as.data.frame(rbind(summary(asw_batch_i)[["si.summary"]]))
  
  asw_cluster_i <- silhouette(as.integer(factor(clusters[subset_id])), dissimilarity_matrix)
  asw_cluster[[i]] <- as.data.frame(rbind(summary(asw_cluster_i)[["si.summary"]]))
}

all_asw_batch <- bind_rows(asw_batch) %>%
  mutate(Cluster = "Batch")

# ASW_batch summary
summary(all_asw_batch$Mean)

all_asw_cluster <- bind_rows(asw_cluster) %>%
  mutate(Cluster = "Cluster")

#ASW_group summary
summary(all_asw_cluster$Mean)

all_asw <- bind_rows(all_asw_batch, all_asw_cluster)

asw_plot <- ggplot(data = all_asw, aes(x = Cluster, y = Mean, fill = Cluster)) +
  geom_violin() +
  geom_boxplot(width = 0.2) +
  labs(x = "") +
  theme(legend.position = "none") +
  scale_fill_few()

asw_plot

all_asw %>%
  group_by(Cluster) %>%
  mutate(Method = method_name) %>%
  write_csv(file = paste0("output/", method_name, "_", dataset_name, "_ASW.csv"))
```

### Calculate LISI Scores

```{r}
# Performed for all cells

lisi <- compute_lisi(integrated_final, data.frame(batch = batches, cluster = clusters), c("batch", "cluster"), perplexity = 40) %>%
  pivot_longer(cols = c("batch", "cluster"), names_to = "type", values_to = "lisi")

# Looking for values close to the number of batches (2)
# Looking for values close to 1

ggplot(data = lisi, aes(x = type, y = lisi, fill = type)) +
  geom_violin() +
  geom_boxplot(width = 0.05)
```

<!-- ### Calculate Adjusted Rand Index -->

<!-- * Compute adjusted rand index (ARI) for a 80% sub-sample of clusters -->
<!--   * ARI is bounded above by 1 (i.e., perfect agreement between two partitions) -->
<!--   * Values **close to 1** indicate the two partition labels are similar -->
<!--   * Values **close to 0** indicate the two partition labels are not similar -->
<!--     * We hope to see lower values for $ARI_{batch}$ since it is indicative of well-mixed batches -->
<!--     * We hope to see higher values for $ARI_{group}$ since it is indicative that distinct cell groups were preserved after batch-mixing -->
<!-- * Repeat `20` times -->
<!-- * Calculate the median ARI from the `20` runs to ensure stability of the measurement -->

<!-- ```{r jive_ari} -->
<!-- ari_repeat <- 20 -->
<!-- npcs <- 30 -->

<!-- ari_batch <- list() -->
<!-- ari_group <- list() -->
<!-- subset_prop <- 0.8 # subsample to 80% of the data -->
<!-- subset_size_total <- floor(length(batch) * subset_prop) -->

<!-- set.seed(1) -->
<!-- for (i in 1:ari_repeat) { -->
<!--   subset_index <-  -->
<!--     data.frame(index = 1:length(batch) , batch = batch) %>% -->
<!--     slice_sample(n = subset_size_total, replace = F) -->

<!--   subset_id <- subset_index %>% -->
<!--     pull(index) -->

<!--   clusters <- kmeans(svd(integrated_final[subset_id, ])$u, centers = length(unique(groups)))$cluster -->

<!--   ari_batch[[i]] <- adjustedRandIndex(factor(batches[subset_id]), clusters[subset_id]) -->
<!--   ari_group[[i]] <- adjustedRandIndex(factor(groups[subset_id]), clusters[subset_id]) -->
<!-- } -->

<!-- all_ari_batch <- data.frame(ari = unlist(ari_batch)) %>% -->
<!--   mutate(cluster = "Batch") -->

<!-- # Lower values indicate well-mixed batches -->
<!-- summary(all_ari_batch$ari) -->

<!-- all_ari_group <- data.frame(ari = unlist(ari_group)) %>% -->
<!--   mutate(cluster = "Group")  -->

<!-- # Higher values indicate cell types were well preserved -->
<!-- summary(all_ari_group$ari) -->

<!-- all_ari <- bind_rows(all_ari_batch, all_ari_group) -->

<!-- ari_plot <- ggplot(data = all_ari, aes(x = cluster, y = ari, fill = cluster)) + -->
<!--   geom_violin() + -->
<!--   geom_boxplot(width = 0.2) -->

<!-- ari_plot -->
<!-- ``` -->

# Session Information

```{r session_info}
sessionInfo()
```
